Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-21 paper.tex
figures
Comments on new experiments

\abstract{The aim of the experiments described in this paper is to
evaluate the use of keyboard and mouse dynamics as an appropriate
non-obtrusive sensory input for an system that is sensitive to the
affective state of its user. Our motivation for starting this research
line has been the lack of tools and methodologies for taking into
account this affective state in learning
environments. In an ideal situation, when instructors have to choose
from  a collection of programming assignments, they should consider the
  student\'s affective state and skills in order to select a learning resource
  with the appropriate level of difficulty. However, neither the data
  or the ability to process it are present in current learning
  management systems. This work tries to address this problem, by
  focused on the capture and pre-processing of data that is going to
  be fed to several machine learning techniques with the objective of classifying the affective
states of users with different levels of expertise
when learning a new programming language. We capture student data from
a web-based platform, a learning management system where students
interact whith programming excersises. We introduce in this paper a
series of pre-processing techniques that are able to convert data from keyboard and mouse dynamics captured from students as
they were coding basic Python programming assignments  into feature
vector, that have been later used for the classification into five affective states:
boredom, frustration, distraction, relaxation and engagement.
The following classification algorithms have been evaluated: k -nearest
neighbors,feed forward neural networks, na\"ive Bayes classifier and the J-48
tree induction , Deep Neural Network, Random Forest, Gradient Boosted Trees and Na\"ive Bayes Kernel ). The best accuracy was around 75\% and was achieved
by the J-48 algorithm. Results show that data gathered from ready-available,
non-obtrusive sensors can be used successfully
as another input to hyebrid classification models in order to predict an
individual\'s affective states, and these in turn can be used to make
students more engaged and thus learn more efficiently.}

@@

\begin{figure}[h!tbp]
\centering
\includegraphics[width=0.45\textwidth]{KMDAffective.png}
\caption{Code Execution and Affect Recognition Architecture.}
\label{fig_process}
\end{figure}

@@

\begin{figure}[h!tb]
\centering
\includegraphics[width=0.45\textwidth]{editorRresult.png}
\caption{Web-based editor used for code evaluation. The lower panel
  indicates (in Spanish) the tests passed successfully and the program
  output.}
\label{fig_editor}
\end{figure}

\begin{figure}[h!tb]
\centering
\includegraphics[width=0.45\textwidth]{KeyDyn.png}
\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
\label{figKD}
\end{figure}

@@

\section{Experiment and results}
\label{sec:exp}
The aim of this research is to evaluate the use of keyboard
and mouse dynamics as an appropriate sensory input for an affective recognition
system in a learning environment for programmers,
who interactively write short programs. An experimental approach was adopted with this aim: Sensory and
quantitative data was collected from learners as they were enrolled in a basic
course of Python programming. This data was then pre-processed using the method
described earlier in Section \ref{sec:method}, and together with the quantitative data obtained from users,
classifiers were trained and validated. The results were then compared
and we arrived to some conclusions regarding the effectiveness of the method. The goal given to
subjects was to solve as many assignments as they could in a period of two weeks.
There was neither time limit nor a minimum amount of time required for a
participant while trying to solve the assignments or complete the tutorial. The
participants were able to stop and resume their interaction with the system at
any time.
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.45\textwidth]{Completed.png}
\includegraphics[width=0.45\textwidth]{classDist.png}
\caption{Completed activities and assignments by learner in ascending order (left) and distribution of emotions reported (right).}
\label{fig_completed}
\end{figure}

@@

\begin{table*}[!t]
\centering
\caption{Performance of every classifier with 10-fold cross validation. Accuracy ($\kappa$) }
\label{tab_performance}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Na\"ive Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% (0.50) & 80.48\% (0.51) & 76.14\% (0.43) & 78.43\% (0.45) \\
    \hline
    Relaxation      & 71.10\% (0.40) & 72.57\% (0.45) & 69.14\% (0.37) & 71.24\% (0.34)\\
    \hline
    Distraction     & 70.33\% (0.43) & 71.00\% (0.43) & 74.00\% (0.50) & 72.52\% (0.39)\\
    \hline
    Frustration     & 74.71\% (0.49) & 73.86\% (0.45) & 72.62\% (0.42) & 78.00\% (0.50)\\
    \hline
    Boredom         & 74.71\% (0.47) & 84.57\% (0.59) & 83.81\% (0.58) & 76.19\% (0.45)\\
    \hline
    \end{tabular}
\end{table*}

\begin{table*}[!t]
\centering
\caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation. }
\label{tab_performance}
    \begin{tabular}{ | l | l | l | }
    \hline
    Classifier   &  Accuracy            & kappa                 \\
    \hline
    Na\"ive Bayes     & 72.52\%  & 0.339  \\
    \hline
    Deep Neural Network     & 78.29\%  & 0.419  \\
    \hline
    Random Forest      & 78.81\%  & 0.360  \\
    \hline
    Gradient Boosted Trees     & 78.95\%  & 0.415  \\
    \hline
    KNN      & 76.19\%  & 0.242  \\
    \hline
    Na\"ive Bayes Kernel     & 80.22\%  & 0.346 \\
    \hline
    J48     & 78.19 \%  & 0.433  \\
    \hline

@@

As part of the data mining process a canonical genetic algorithm was used for feature
selection with the following parameters: population size 700, 30 generations with
tournament selection, tournament size was 25\% of population size. A uniform crossover
with 0.5 probability and mutation rate of 1/39, and a minimum feature size of 5.
In the end, the subset for the frustration classifier includes 11 features, and the subset for the boredom classifier consists of
13 features.
% It's not clear what's the fitness function here or what it really does. Maybe we could expand this part of the work, test different evolutionary algorithms and write a paper on feature selection using evolutionary algorithms for EvoStar - JJ

@@

>>>>

\abstract{The aim of the experiments described in this paper is to
evaluate the use of keyboard and mouse dynamics as an appropriate
non-obtrusive sensory input for an system that is sensitive to the
affective state of its user. Our motivation for starting this research
line has been the lack of tools and methodologies for taking into
account this affective state in learning
environments. In an ideal situation, when instructors have to choose
from  a collection of programming assignments, they should consider the
  student\'s affective state and skills in order to select a learning resource
  with the appropriate level of difficulty. However, neither the data
  or the ability to process it are present in current learning
  management systems. This work tries to address this problem, by
  focused on the capture and pre-processing of data that is going to
  be fed to several machine learning techniques with the objective of classifying the affective
states of users with different levels of expertise
when learning a new programming language. We capture student data from
a web-based platform, a learning management system where students
interact whith programming excersises. We introduce in this paper a
series of pre-processing techniques that are able to convert data from keyboard and mouse dynamics captured from students as
they were coding basic Python programming assignments  into feature
vector, that have been later used for the classification into five affective states:
boredom, frustration, distraction, relaxation and engagement.
The following classification algorithms have been evaluated: k-nearest
neighbors,feed forward neural networks, na\"ive Bayes classifier, J-48
tree induction algorithm, deep learning, random forest, gradient boosted trees and na\"ive Bayes Kernel). The best accuracy was around 78\% and was achieved
by the tree induction algorithms. Results show that data gathered from ready-available,
non-obtrusive sensors can be used successfully
as another input to hyebrid classification models in order to predict an
individual\'s affective states, and these in turn can be used to make
students more engaged and thus learn more efficiently.}

@@

\begin{figure*}[h!tbp]
\centering
\includegraphics[width=0.80\textwidth]{KMDAffective.png}
\caption{Code Execution and Affect Recognition Architecture.}
\label{fig_process}
\end{figure*}

@@

\begin{figure*}[h!tb]
\centering
\includegraphics[width=0.60\textwidth]{editorRresult.png}
\caption{Web-based editor used for code evaluation. The lower panel
  indicates (in Spanish) the tests passed successfully and the program
  output \cite{ijcci17}.}
\label{fig_editor}
\end{figure*}

\begin{figure*}[h!tb]
\centering
\includegraphics[width=0.70\textwidth]{KeyDyn.png}
\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
\label{figKD}
\end{figure*}

@@

\section{Experiment and results}
\label{sec:exp}
The aim of this research is to evaluate the use of keyboard
and mouse dynamics as an appropriate sensory input for an affective recognition
system in a learning environment for programmers,
who interactively write short programs. An experimental approach was adopted with this aim: Sensory and
quantitative data was collected from learners as they were enrolled in a basic
course of Python programming. This data was then pre-processed using the method
described earlier in Section \ref{sec:method}, and together with the quantitative data obtained from users,
classifiers were trained and validated. The results were then compared
and we arrived to some conclusions regarding the effectiveness of the method. The goal given to
subjects was to solve as many assignments as they could in a period of two weeks.
There was neither time limit nor a minimum amount of time required for a
participant while trying to solve the assignments or complete the tutorial. The
participants were able to stop and resume their interaction with the system at
any time.
%
\begin{figure*}[!t]
\centering
\includegraphics[width=0.48\textwidth]{Completed.png}
\includegraphics[width=0.48\textwidth]{classDist.png}
\caption{Completed activities and assignments by learner in ascending order (left) and distribution of emotions reported (right) \cite{ijcci17}.}
\label{fig_completed}
\end{figure*}

@@

\begin{table*}[!t]
\centering
\caption{Performance of every classifier with 10-fold cross validation \cite{ijcci17}. Accuracy ($\kappa$) }
\label{tab_performance}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Na\"ive Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% (0.50) & 80.48\% (0.51) & 76.14\% (0.43) & 78.43\% (0.45) \\
    \hline
    Relaxation      & 71.10\% (0.40) & 72.57\% (0.45) & 69.14\% (0.37) & 71.24\% (0.34)\\
    \hline
    Distraction     & 70.33\% (0.43) & 71.00\% (0.43) & 74.00\% (0.50) & 72.52\% (0.39)\\
    \hline
    Frustration     & 74.71\% (0.49) & 73.86\% (0.45) & 72.62\% (0.42) & 78.00\% (0.50)\\
    \hline
    Boredom         & 74.71\% (0.47) & 84.57\% (0.59) & 83.81\% (0.58) & 76.19\% (0.45)\\
    \hline
    \end{tabular}
\end{table*}

\begin{table*}[!t]
\centering
\caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation. }
\label{tab_perf_flow}
    \begin{tabular}{ | l | l | l | }
    \hline
    Classifier   &  Accuracy            & $\kappa$                 \\
    \hline
    Na\"ive Bayes     & 72.52\%  & 0.339  \\
    \hline
    Deep Neural Network     & 78.29\%  & 0.419  \\
    \hline
    Random Forest      & 78.81\%  & 0.360  \\
    \hline
    Gradient Boosted Trees     & 78.95\%  & 0.415  \\
    \hline
    KNN      & 76.19\%  & 0.242  \\
    \hline
    Na\"ive Bayes Kernel     & 80.22\%  & 0.346 \\
    \hline
    J48     & 78.19 \%  & 0.433  \\
    \hline

@@

As part of the data mining process a canonical genetic algorithm was used for feature
selection with the following parameters: population size 700, 30 generations with
tournament selection, tournament size was 25\% of population size. A uniform crossover
with 0.5 probability and mutation rate of 1/39, and a minimum feature size of 5.
In the end, the subset for the frustration classifier includes only 11 features, and 
the subset for the boredom classifier consists of only 13 features.

% It's not clear what's the fitness function here or what it really does. Maybe we could expand this part of the work, test different evolutionary algorithms and write a paper on feature selection using evolutionary algorithms for EvoStar - JJ

@@

Table \ref{tab_perf_flow} shows the performance for the additional classifiers
evaluated specifically for the Flow/engaged affective state. The goal was to
increase the performance achieved earlier. Table 1shows the performance for the
additional classifiers evaluated specifically for the Flow/engaged affective
state. The goal was to increase the performance achieved earlier.  Again a ten-
fold cross-validation was conducted. In this case, instead of using an
evolutionary algorithm to conduct feature selection, a vector of weights was
optimized. The feature-weight vector assignes a weight to each feature. The
parameters used for the evolutionary optimization algorithm are the following:
Population size= 10, generations= 30, selection= tournament, tournament size=
0.25 (fraction of the population),  initialize probability=0.5,  mutation
probability = 1/number of attributes,  crossover probability=0.5. Parameters
used for the Random Forest algorithm: number of trees=10, criterion=gain_ratio,
maximal depth=20, confidence=0.25, minimal gain=0.1, minimal leaf size=2.  The
parameters used for are the J48 are the following:  confidence=0.79, minimal
leaf size=2. For the kNN algorithm, the value k=4 was used. For gradient boosted
trees: number of trees= 20,  maximal depth=5, minimum rows=10, minimum split
improvement=0.0, number of bins=20, learning rate=0.1, sample rate=1.0.  Bayes
Kernel:  laplace correction=True, estimation mode=greedy, minimum bandwidth=0.1,
number of kernels=10.  Finally for deep learning: activation=Rectifier, hidden
layer sizes= [50, 50], epochs=10. Again the accuracy and $\kappa$ values of the
algorithms show the difficulty of having classes that are not uniformly
distributed, with several algorithms reaching an acceptable accuracy but with a
low value of  $\kappa$, this is an indication of algorithms classifying almost
everything as the dominant class. The results of the tree induction algorithms
are better when we consider both accuracy and  $\kappa$ values. In the case of
kernel-based or deep learning methods they provide competitive results but at a
higher computational cost.
